---
title: "eval_EDA"
output: html_document
---
This will serve as a place to explore visualisations and analysis from the [U.S. Variant Nowcast Hub](https://github.com/reichlab/variant-nowcast-hub).
```{r setup, message = FALSE, warning = FALSE, results = 'hide'}
library(ggplot2)
library(dplyr)
library(hubData)
library(readr)
library(arrow)
library(scoringutils)
```

Test out loading in some model outputs for a single nowcast date 
```{r}
this_nowcast_date <- "2025-04-23"
bucket_name <- "covid-variant-nowcast-hub"
state<- "MA"
hub_bucket <- s3_bucket(bucket_name)
hub_con <- hubData::connect_hub(hub_bucket, file_format = "parquet", skip_checks = TRUE)
MA_data<- hub_con |>
  dplyr::filter(location == state, 
                clade != "recombinant",
                nowcast_date == this_nowcast_date) |>
  hubData::collect_hub() |>
  dplyr::select(model_id, nowcast_date, target_date, location, clade, value, 
                output_type, output_type_id)
```
Load in some target data for a single nowcast date (e.g. April 23)
```{r}
target_data <- arrow::read_parquet(glue::glue("https://github.com/reichlab/variant-nowcast-hub/raw/refs/heads/adding-scores/target-data/oracle-output/nowcast_date={this_nowcast_date}/oracle.parquet")) |>
  filter(location == state)

```

Load in some scores (this will be for all nowcast dates that we can scores)
```{r}
scores <- read_tsv("https://raw.githubusercontent.com/reichlab/variant-nowcast-hub/refs/heads/adding-scores/auxiliary-data/scores/scores_2025-09-02.tsv") |>
  filter(location == state, nowcast_date == this_nowcast_date)
```
Plot eval data and model outputs (latent proportions) for this nowcast date
```{r}
clades_to_vis <- c("24F", "24A", "25A", "24E")
MA_model_outputs <- MA_data |> 
  dplyr::filter(clade %in% clades_to_vis) |>
  mutate(output_type_id_clade = glue::glue("{output_type_id}-{clade}"))
n_by_day <- target_data_2025_04_23 |>
  group_by(target_date) |>
  summarise(n_seq = sum(oracle_value))
eval_data <- target_data_2025_04_23 |>
   filter(clade %in% clades_to_vis) |>
  left_join(n_by_day, by = "target_date") |>
  ungroup() |>
  mutate(obs_freq = oracle_value/n_seq)

ggplot() +
  geom_line(data = MA_model_outputs  |>
              filter(output_type == "mean"),
            aes(x = target_date, y = value, color = clade)) +
   geom_line(data = MA_model_outputs |>
              filter(output_type == "sample"),
            aes(x = target_date, y = value, color = clade,
                group = output_type_id_clade),
            size = 0.2, alpha = 0.2) +
  geom_point(data = eval_data,
             aes(x = target_date, y = obs_freq, color = clade,
                 size = n_seq)) +
  facet_wrap(~model_id) +
  theme_bw() 
```

However, we really want to see the data against the prediction intervals 
using the multinomial sampling procedure-- we can do this by expanding the 
outputs to include samples from the predictive distribution. 
```{r}
MA_preds <- MA_data |>
  
  left_join(n_by_day, by = "target_date") |>
  group_by(model_id, target_date, location, nowcast_date, output_type_id) |>
  summarise(
     n_seq = first(n_seq),
      samples = list(lapply(1:100, function(i) {
      counts <- as.vector(rmultinom(1, size = first(n_seq), prob = value))
      tibble(clade = clade, count = counts, sample_id = i)
    })),
    .groups = "drop"
  ) |>
  tidyr::unnest(samples) |>
  tidyr::unnest(samples)
```
Something not looking right... 
```{r}
ggplot() +
  geom_line(data = MA_model_outputs  |>
              filter(output_type == "mean"),
            aes(x = target_date, y = value, color = clade)) +
   geom_line(data = MA_preds |>
               mutate(sample_id_clade = glue::glue("{clade}-{sample_id}-{output_type_id}")),
            aes(x = target_date, y = count/n_seq, color = clade,
                group = sample_id_clade),
            size = 0.2, alpha = 0.2) +
  geom_point(data = eval_data,
             aes(x = target_date, y = obs_freq, color = clade,
                 size = n_seq)) +
  facet_wrap(~model_id) +
  theme_bw() 
```
Look at scores
```{r}
ggplot(scores) +
  geom_line(aes(x = target_date, y = energy)) +
  facet_wrap(~model_id) + theme_bw() +ggtitle("Energy")

ggplot(scores) +
  geom_line(aes(x = target_date, y = brier)) +
  facet_wrap(~model_id) + theme_bw() +ggtitle("Brier")

scores_summary <- scores |>
  group_by(model_id) |>
  summarise(mean_energy = mean(energy, na.rm = TRUE),
            mean_brier = mean(brier, na.rm = TRUE)
  )

ggplot(scores_summary) + 
  geom_bar(aes(x = model_id, y = mean_energy,
               fill = model_id),
           stat = "identity", position = "dodge") +
  theme_bw()  + ggtitle("Energy")

ggplot(scores_summary) + 
  geom_bar(aes(x = model_id, y = mean_brier,
               fill = model_id),
           stat = "identity", position = "dodge") +
  theme_bw()  + ggtitle("Brier")
```
Compute scaled relative skill
```{r}
scores2 <- scores |>
  rename(energy_score = energy,
         brier_score= brier,
         model = model_id) |>
  select(location, target_date, nowcast_date, model, energy_score,
         brier_score) |>
  data.table::as.data.table()
  class(scores2) <- c("scores", class(scores2))
  scores_su <- data.table::setattr(scores2,
                                   "metrics",
                                   c("brier_score", "energy_score"))
rel_skill_es_t <- scores_su |>
    scoringutils::add_relative_skill(metric = "energy_score",
                                   baseline = "Hub-baseline",
                                   by = "target_date")
rel_skill_bs_t <- scores_su |>
  scoringutils::add_relative_skill(metric = "brier_score",
                                   baseline = "Hub-baseline",
                                   by = "target_date")
ggplot(rel_skill_es_t) +
  geom_line(aes(x = target_date, y = energy_score_scaled_relative_skill, color = model))

ggplot(rel_skill_bs) +
  geom_line(aes(x = target_date, y = brier_score_scaled_relative_skill, color = model))


```

Repeat but overall
```{r}
rel_skill_es <- scores_su |>
    scoringutils::add_relative_skill(metric = "energy_score",
                                   baseline = "Hub-baseline") |>
  group_by(model) |> 
  summarise(mean_rel_skill = mean(energy_score_relative_skill, na.rm = TRUE),
            mean_rel_scaled_skill = mean(energy_score_scaled_relative_skill, 
                                         na.rm = TRUE))
rel_skill_bs <- scores_su |>
  scoringutils::add_relative_skill(metric = "brier_score",
                                   baseline = "Hub-baseline") |>
  group_by(model) |>
  summarise(mean_rel_skill = mean(brier_score_relative_skill, na.rm = TRUE),
            mean_rel_scaled_skill = mean(brier_score_scaled_relative_skill, 
                                         na.rm = TRUE))

ggplot(rel_skill_bs) +
  geom_bar(aes(x = model, y = mean_rel_scaled_skill, fill = model),
           stat = "identity", position = "dodge") +
  ggtitle("Brier score")
ggplot(rel_skill_es) +
  geom_bar(aes(x = model, y = mean_rel_scaled_skill, fill = model),
           stat = "identity", position = "dodge") +
  ggtitle("Energy score")
```

