---
title: "Try out existing baseline code"
output: html_document
date: "2025-01-21"
---

This code tests out using code from the RESPINOW-Hub baseline model from 
https://github.com/KITmetricslab/RESPINOW-Hub/blob/main/code/baseline/run_baseline.R

A first step of the baseline package will likely be to use this/build off of 
this code to create the package and then modify as needed. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(epinowcast)
library(zoo)
library(dplyr)
library(ggplot2)
library(data.table)
library(here)
```

## Load in the baseline functions
Eventually this will be installed via an R package but for now just source them
from a local clone of the repo in an adjacent folder
```{r}
source("../../RESPINOW-Hub/code/baseline/functions.R")
source("../../RESPINOW-Hub/respinow_viz/functions.R")

# Also load in custom functions we created here (mostly modifying RESPINOW functions)
source("preprocess_rt.R") # removed any hard coding of the reporting triangle being in weeks
source("data_as_of_new.R") # fewer assumptions about columns (year and week not necessary) + 
# doesn't assume that the reporting triangle is in weeks and needs an additional delay to be added in days. 
# will therefore return in the same units as input reporting triangle (so this might be weeks) 
source("compute_exp.R")
```

Load in the data we will use. Here we're going to use the epinowcast package 
data as a starting point. We'll exclude a portion of it that is complete so 
that we can test the nowcasts. The reference dates and report dates run from 
April to October of 2021. We will truncate report dates after August, this way
we have some dates with complete observations (April - May), and some dates
with partial observations. We can then estimate the final observations for May
to August and evaluate using the full data, since by October all of the 
partially observed data will have been fully reported. 
```{r}
# start with just one age group, one pathogen, one location... 
full_data_long <- epinowcast::germany_covid19_hosp[location == "DE"][age_group == "00+"]

forecast_date <- "2021-07-15"
training_data_long <- full_data_long |>
  enw_filter_report_dates(latest_date = forecast_date)
```

Use epinowcast functions to get the reporting triangle
```{r}
max_delay_enw <- 80
pobs <- enw_preprocess_data(training_data_long, max_delay = max_delay_enw)
triangle <- pobs$reporting_triangle[[1]]
```

## Use RESPINOW functions to compute the nowcast 
We will use the `compute_nowcast()` function, but first we need to 
reformat the epinowcast data to work with the RESPINOW functions
```{r}

# Reformat reporting triangle
max_delay <- ncol(triangle) - 3 # 2 metadata columns + 0 
triangle_df <- as.data.frame(triangle)
cols_to_prefix <- as.character(c(0:(max_delay)))
# Remove the prefix, since we will modify preprocess_reporting_triangle to
# take in a generic reporting triangle with no specified units 
colnames(triangle_df)[colnames(triangle_df) %in% cols_to_prefix] <- paste0("value_", cols_to_prefix)
triangle_df <- triangle_df |> 
  mutate(age_group = "00+", location = "DE") |>
  rename(date = reference_date) |>
  select(location, age_group, date, c(paste0("value_", cols_to_prefix)))


# We will start by not using the borrowing of delays or dispersion functionality,
# but later we can test out that functionality by passing in the data for 
# all the age groups and applying it to a specific age group. 

# Other parameters we will need:

nowcast <- compute_nc(observed = triangle_df, # Reporting triangle, formatted as compute_nowcast expects 
                              location = "DE",
                              age_group = "00+",
                              forecast_date = lubridate::ymd(forecast_date),
                              borrow_delays = FALSE,
                              borrow_dispersion = FALSE,
                              observed2 = NULL,
                              location2 = NULL,
                              age_group2 = NULL,
                              type = "additions",
                              max_delay = max_delay,
                              n_history_expectations = nrow(triangle_df), # number of reference dates to use in estimation of delay?
                              n_history_dispersion = nrow(triangle_df), # number of reference dates to use in estimation of dispersion 
                              )

```

